Introducing ICBe: Very High Recall and Precision Event Extraction from Narratives about International Crises

### 2022-09-29 replication_paper/political_analysis_draft/ICBe_political_analysis_submission_rmd.Rmd
spell-check

abstract: |
  How do international crises unfold? We conceptualize of international relations as a strategic chess game between adversaries and develop a systematic way to measure pieces, moves, and gambits accurately and consistently over a hundred years of history. We introduce a new ontology and dataset of international events called ICBe based on a very high-quality corpus of narratives from the International Crisis Behavior (ICB) Project. We demonstrate that ICBe has higher coverage, recall, and precision than existing state of the art datasets and conduct two detailed case studies of the Cuban Missile Crisis (1962) and Crimea-Donbas Crisis (2014). We further introduce two new event visualizations (event icongraphy and crisis maps), an automated benchmark for measuring event recall using natural language processing (sythnetic narratives), and an ontology reconstruction task for objectively measuring event precision. We make the data, supplementary appendix, replication material, and visualizations of every historical episode available at a companion website www.crisisevents.org.

@@

If we could record every important interaction between countries in all of diplomacy, military conflict, and international political economy, how much unique information would this chronicle amount to, and how surprised would we be to see something new? In other words, what is the entropy of international relations? This record could in principle be unbounded, but the central conceit of social science is that there are structural regularities that limit what actors can do, their best options, and even which actors are likely to survive [@brecherInternationalStudiesTwentieth1999; @reiterShouldWeLeave2015]. If so, then these events can be systematically measured, and accordingly, massive effort is expended in social science attempting to record these regularities.^[See work on crises [@brecherCrisesWorldPolitics1982; @beardsleyInternationalCrisisBehavior2020], militarized disputes [@palmerMID5Dataset20112021; @giblerInternationalConflicts181620102018; @maozDyadicMilitarizedInterstate2019], wars [@sarkeesResortWar181620072010; @reiterRevisedLookInterstate2016], organized violence [@ralphsundbergUCDPGEDCodebook2016; @petterssonOrganizedViolence19892018], political violence [@raleighIntroducingACLEDArmed2010], sanctions [@felbermayrGlobalSanctionsData2020], trade [@barariDemocracyTradePolicy], and international agreements [@kinneDefenseCooperationAgreement2020; @owsiakInternationalBorderAgreements2018; @vabulasCooperationAutonomyBuilding2021], dispute resolution [@vabulasCooperationAutonomyBuilding2021; @frederickIssueCorrelatesWar2017], and diplomacy [@moyerWhatAreDrivers2020; @sechserMilitarizedCompellentThreats2011].] Thanks to improvements in natural language processing, more open-ended efforts have begun to capture entire unstructured streams of events from international news reports.^[See @liComprehensiveSurveySchemabased2021; @haltermanExtractingPoliticalEvents2020; @brandtPhoenixRealTimeEvent2018; @boscheeICEWSCodedEvent2015; @hegreIntroducingUCDPCandidate2020; @grantOUEventData2017. On event-extraction from images and social-media see @zhangCASMDeepLearningApproach2019 and @steinert-threlkeldFutureEventData2019.] How close these efforts are to accurately measuring all or even most of what is essential in international relations is an open empirical question, one for which we provide new evidence here.

Our contribution is a high coverage ontology and event dataset for key historical episodes in 20th and 21st-century international relations. We develop a large, flexible ontology of international events with the help of both human coders and natural language processing. We apply it sentence-by-sentence to an unusually high-quality corpus of historical narratives of international crises [@brecherInternationalStudiesTwentieth1999; @brecherCrisisEscalationWar2000; @wilkenfeldInterstateCrisesViolence2000; @jamesWhatWeKnow2019; @iakhnisCrisesWorldPolitics2019]. The result is a new lower bound estimate of how much actually happens between states during pivotal historical episodes. We then develop several methods for objectively gauging how well these event codings reconstruct the information contained in the original narrative. We conclude by benchmarking our event codings against several current state-of-the-art event data collection efforts. We find that existing systems produce sequences of events that do not contain enough information to reconstruct the underlying historical episode. The underlying fine-grained variation in international affairs is unrecognizable through the lens of current quantification efforts.

This is a measurement paper that makes the following argument --- there is a real-world unobserved latent concept known as international relations, we propose a method for systematically measuring it, we successfully apply this method producing a new large scale set of measurements, those measurements exhibit several desirable kinds of internal and external validity, and those measurements out-perform other existing approaches. The article organizes that argument into eight sections: task definition; corpus; priors/existing state of the art; ICBe coding process; internal consistency; case study selection; recall; and precision. A final section concludes.

@@

We consider the measurement task of abstracting discrete events about a historical episode in international relations. The easiest way to convey the task is with an example. Figure 1 shows a narrative account of the Cuban Missile Crisis (1962) alongside a mapping from each natural language sentence to discrete machine readable abstractive events. Formally, a historical episode, $H$, is demarcated by a period of time $[T_{start}, T_{end}] \in T$, a set of Players $p \in P$, and a set of behaviors they undertook during that time $b \in B$. International Relations, $IR$, is the system of regularities that govern the strategic interactions that world actors make during a historical episode, given their available options, preferences, beliefs, and expectations of choices made by others. We observe neither $H$ nor $IR$ directly. Rather the Historical Record, $HR$, produces documents $d \in D$ containing some relevant and true (as well as irrelevant and untrue) information about behaviors that were undertaken recorded in the form of unstructured natural language text. The task is to combine informative priors about $IR$ with an unstructured corpus $D$ to produce a series of structured discrete events, $e \in E$, that have high coverage, precision, and recall over what actually took place in history, $H$.

@@

temp <- ICBe_events_agreed_markdown %>%
        filter(crisno==196) %>%
        group_by(crisno, sentence_number_int_aligned) %>%
          filter(!duplicated(value_markdown %>% trimws() )) %>% #note there are dupe markdowns for some reason
        ungroup() %>%
        mutate(banding=as.numeric(as.factor(sentence_number_int_aligned)) %% 2)
temp$is_subsstring <- NA
for(i in 1:nrow(temp)){
  temp$is_subsstring[i] <-   stringr::str_detect(temp$value_markdown %>% tolower() %>% stringr::str_replace_all('[^a-z0-9]',"") ,
                                                                                pattern = stringr::fixed(temp$value_markdown[i] %>% tolower() %>% stringr::str_replace_all('[^a-z0-9]',"") )
                                                                              ) %>% replace_na(0) %>%  sum()
}
temp <-temp %>%
        mutate(value_markdown = ifelse(is_subsstring<=1,value_markdown,'')) %>%
        group_by(sentence_number_int_aligned) %>%
          filter(value_markdown!='' | length(value_markdown)==1) %>% # also added below to handle sentence 6
        ungroup() %>%
        mutate(crisno= ifelse(!is.na(lag(sentence_span_text)) & sentence_span_text==lag(sentence_span_text), "", crisno)) %>%
        mutate(sentence_number_int_aligned= ifelse(!is.na(lag(sentence_span_text)) &  sentence_span_text==lag(sentence_span_text), "", sentence_number_int_aligned)) %>%
        mutate(sentence_span_text= ifelse(!is.na(lag(sentence_span_text)) &  sentence_span_text==lag(sentence_span_text), "", sentence_span_text)) %>% 
            filter(value_markdown!='' | length(value_markdown)==1) 
      #C=crisno, 
      #head(100) %>%
ft_cuba_precision <- temp %>%
                     dplyr::select(S=sentence_number_int_aligned, ICB=sentence_span_text, ICBe=value_markdown)  %>%
                      flextable::as_flextable() %>%
        flextable::set_header_labels(values =
        list(
              S = "S",
              ICB = "Natural Language Sentences (ICB Corpus)",
              ICBe = "Machine Readable Events (ICBe)"
             ) ) %>%

@@

Next we draw informative prior beliefs about the underlying process of IR that we expect to govern behavior during historical episodes and their conversion to the historical record. We organize our prior beliefs along two overarching axes, summarized in detail by Table 1.

@@

We base our informed priors primarily on two sources of information. The first is the extensive existing measurement efforts of IR which we provide citations to alongside each concept. Second, we performed preliminary natural language processing of the corpus and identified named entities and behaviors mentioned in the text. Verbs were matched to the most likely definition found in Wordnet [@millerWordNetLexicalDatabase1995], tallied, and then aggregated into a smaller number hypernyms balancing conceptual detail and manageable sparsity for human coding (SI Appendix 1.2).

The second axis (columns) compares the very high ontological coverage of ICBe to existing state of the art systems in production and with global coverage. They begin with our contribution ICBe, alongside other event-level datasets including CAMEO dictionary lookup based systems (Historical Phoenix [@althausClineCenterHistorical2019]; ICEWS [@boscheeICEWSCodedEvent2015; @hegreIntroducingUCDPCandidate2020]; Terrier [@grantOUEventData2017]), the Militarized Interstate Disputes Incidents dataset, and the UCDP-GED dataset [@ralphsundbergUCDPGEDCodebook2016; @petterssonOrganizedViolence19892018; @sundbergIntroducingUCDPGeoreferenced2013].^[Additional relevant but dated or too small of an overlap in domain include BCOW [@lengMilitarizedInterstateCrises1988], WEIS [@mcclellandWorldEventInteraction1978], CREON [@hermannComparativeResearchEvents1984], CASCON [@bloomfieldCASCONIIIComputeraided1989], SHERFACS [@shermanSHERFACSCrossParadigmHierarchical2000], Real-Time Phoenix [@brandtPhoenixRealTimeEvent2018], and COfEE [@balaliCOfEEComprehensiveOntology2021] (see histories in @merrittMeasuringEventsInternational1994 and @schrodtTwentyYearsKansas2006).] The final set of columns compares episode-level datasets beginning with the original ICB project [@brecherInternationalCrisisBehavior; @brecherCrisesWorldPolitics1982; @beardsleyInternationalCrisisBehavior2020]; the Militarized Interstate Disputes dataset [@palmerMID5Dataset20112021; @giblerInternationalConflicts181620102018;@braithwaiteMIDLOCIntroducingMilitarized2010;@braithwaiteCodebookMilitarizedInterstate2009], and the Correlates of War [@sarkeesResortWar181620072010]. With the exception of large scale CAMEO dictionary based systems, the existing state of the art quantitative datasets ignore the vast majority of the information content found in international relations.^[See [@balaliCOfEEComprehensiveOntology2021] for a recent review of ontological depth and availability of Gold Standard example text.]

@@

The ICBe ontology follows a hierarchical design philosophy where a smaller number of significant decisions are made early on and then progressively refined into more specific details [@brustIntegratingDomainKnowledge2020].^[This process quickly focuses the coder on a smaller number of relevant options while also allowing them to apply multiple tags if the sentence explicitly includes more than one or there is insufficient evidence to choose only one tag. The guided coding process also allows for the possibility that earlier coarse decisions have less error than later fine-grained decisions.] Each coder was instructed to first thoroughly read the full crisis narrative and then presented with a custom graphical user interface (SI Appendix 2.1). Coders then proceeded sentence by sentence, choosing the number of events (0-3) that occurred, the highest behavior (thought, speech, or activity), a set of players ($P$), whether the means were primarily armed or unarmed, whether there was an increase or decrease in aggression (uncooperative/escalating or cooperative/de-escalating), and finally one or more non-mutually exclusive specific activities. Some additional details like location and timing information was always collected while other details were only collected if appropriate, e.g. force size, fatalities, domains, units, etc. A unique feature of the ontology is that thought, speech, and do behaviors can be nested into combinations, e.g. an offer for the U.S.S.R. to remove missiles from Cuba in exchange for the U.S. removing missiles from Turkey. Through compounding, the ontology can capture what players were said to have known, learned, or said about other specific fully described actions.

@@

We attribute the remaining disagreement primarily to three sources. First, we required coders to rate and justify their confidence in the coding. They reported low confidence for 20% of sentences; 45% of those were due to a mismatch between the ontology and the text ("survey doesn't fit event") and 46% were from a lack of information or confused writing in the source text (40% "more knowledge needed", 6% "confusing sentence"). Observed disagreement varied predictably with self reported confidence (SI Appendix 2.4). Second, as intended, agreement is higher (75-80%) for questions with fewer options near the root of the ontology compared to agreement for questions near the leafs of the ontology (50%-60%). Third, individual coders exhibiting nontrivial coding styles, e.g. some more expressive applying many tags per concept while others focused on only the single best match. We further observed unintended synonymity, e.g. the same information can be framed as either a threat to do something or a promise not to do something.

@@

The remaining two qualities we seek to measure are recall and precision of ICBe events in absolute terms and relative to other existing systems. We provide full ICB narratives, ICBe coding in an easy to read icongraphic form, and a wide range of visualizations for every case on the companion website. In this paper, we focus on two deep case studies. The first is the Cuban Missile Crisis (Figure 1) which took place primarily in the second half of 1962, involved the United States, the Soviet Union, and Cuba, and is widely known for bringing the world to the brink of nuclear war (hereafter Cuban Missiles). The second is the Crimea-Donbas Crisis (SI Appendix 4.1) which took place primarily in 2014, involved Russia, Ukraine, and NATO, and within a decade spiraled into a full scale invasion (herafter Crimea-Donbas). Both cases involve a superpower in crisis with a neighbor, initiated by a change from a friendly to hostile regime, with implications for economic and military security for the superpower, risked full scale invasion, and eventually invited intervention by opposing superpowers. We choose these cases because they are substantively significant to 20th and 21st century international relations, widely known across scientific disciplines and popular culture, and are sufficiently brief to evaluate in depth.

@@

Recall measures the share of desired information recovered by a sequence of coded events, $Pr(E|H)$, and is poorly defined for historical episodes. First, there is no genuine ground truth about what occurred, only surviving texts about it. Second, there is no \textit{a priori} guide to what information is necessary detail and what is ignorable trivia. History suffers from what is known as the Coastline Paradox [@mandelbrotFractalGeometryNature1983] --- it has a fractal dimension greater than one such that the more you zoom in the more detail you will find about individual events and in between every two discrete events. The ICBe ontology is a proposal about what information is important, but we need an independent benchmark to evaluate whether that proposal is a good one that allows for comparing proposals from event projects that had different goals. We need a yardstick for history.

Our strategy for dealing with both problems is a plausibly objective yardstick called a synthetic historical narrative. For both case studies, we collect a large diverse corpus of narratives spanning timelines, encyclopedia entries, journal articles, news reports, websites, and government documents. Using natural language processing (fully described in SI Appendix 3.1), we identify details that appear across multiple accounts. The more accounts that mention a detail, the more central it is to understanding the true historical episode. The theoretical motivation is that authors face word limits which force them to pick and choose which details to include, and they choose details which serve the specific context of the document they are producing. With a sufficiently large and diverse corpus of documents, we can vary the context while holding the overall episode constant and see which details tend to be invariant to context. Intuitively, a high quality event dataset should have high recall for context invariant details both because of their broader relevance and also because they are easier to find in source material.

Synthetic historical narratives for Cuban Missiles (51 events drawn from 2020 documents) and Crimea-Donbas (30 events drawn from 971 documents) appear in Figure 2. Each row represents a detail which appeared in at least five documents along with an approximate start date, a hand written summary, the number of documents it was mentioned in, and whether it could be identified in the text of our ICB corpus, in our ICBe events, or any of the competing systems.

From them, we draw several stylized facts. First, there is substantial variation in which details any one document will choose to include. Our ground truth ICB narratives included 17/51 and 23/30 of the events from the synthetic narrative, while including other details that are not in the synthetic narrative. Second, mentions of a detail across accounts is exponentially distributed with context invariant details appearing dozens to hundreds of times more than context dependent details. Third, crisis start and stop dates are arbitrary and the historical record points to many precursor events as necessary detail for understanding later events, e.g. the U.S. was in a \textit{de facto} grey scale war with Cuba before it invited Soviet military protection [@cormacGreyNewBlack2018] and Ukraine provided several security guarantees to Russia that were potentially undone, e.g. a long term lease on naval facilities in Crimea. Fourth, we find variation between the two cases. Cuban Missiles has a cleaner canonical end with the Soviets agreeing to withdraw missiles while Crimea-Donbas meekly ends with a second cease fire agreement (Minsk II) but continued fighting. The canonical narrative of Cuban Missile also includes high level previously classified details, while the more recent Crimea-Donbas case reflects primarily public reporting.

We find substantive variation in recall across systems. Recall for each increases in the number of document mentions which is an important sign of validity for both them and our benchmark. The one outlier is Phoenix which is so noisy that it's flat to decreasing in mentions. The two episode level datasets have very low coverage of contextual details. The two other dictionary systems ICEWs and Terrier have high coverage, with ICEWs outperforming Terrier. ICBe strictly dominates all of the systems but ICEWs in recall though we note that the small sample sizes mean these systems should be considered statistically indistinguishable. Importantly our corpus of ICB narratives has very high recall of frequently mentioned details giving us confidence in how those summaries were constructed, and ICBe lags only slightly behind showing that it left very little additional information on the table.

@@

      add_header_row(values = c("Cuban Missile Crisis (1962)", "", "Crimea-Donbass (2014)"), colwidths = c(kleft, 1,kright) ) %>%

@@

The other side of event measurement is precision, the degree to which a sequence of events correctly and usefully describes the information in history, $Pr(H|E)$. It does little good to recall a historical event but too vaguely (e.g. MIDs describes the Cuban Missile crisis as a blockade, a show of force, and a stalemate) or with too much error (e.g. ICEWS records 263 "Detonate Nuclear Weapons" events between 1995-2019) to be useful for downstream applications. ICBe's ontology and coding system is designed to strike a balance so that the most important information is recovered accurately but also abstracted to a level that is still useful and interpretable. You should be able to lay out events of a crisis on a timeline, as in Figure 3, and read off the macro structure of an episode from each individual move. We call this visualization a crisis map, a directed graph intersected with a timeline, and provide crisis maps for every event dataset for each case study in SI Appendix 4.3 and 4.4 and for all crises on the companion website.

@@

We further want to verify individual event codings, which we can do in the case of ICBe because each event is mapped to a specific span of text. We develop the iconography system for presenting event codings as coherent statements that can be compared side by side to the original source narrative as for Cuban Missiles (Figure 1), Crimea-Donbas (SI Appendix 4.1), and for every case on the companion website. We further provide a stratified sample of event codings alongside their source text (SI Appendix 4.2).

>>>>

abstract: |
  How do international crises unfold? We conceptualize international relations as a strategic chess game between adversaries and develop a systematic way to measure pieces, moves, and gambits accurately and consistently over a hundred years of history. We introduce a new ontology and dataset of international events called ICBe based on a very high-quality corpus of narratives from the International Crisis Behavior (ICB) Project. We demonstrate that ICBe has higher coverage, recall, and precision than existing state of the art datasets and conduct two detailed case studies of the Cuban Missile Crisis (1962) and the Crimea-Donbas Crisis (2014). We further introduce two new event visualizations (event iconography and crisis maps), an automated benchmark for measuring event recall using natural language processing (synthetic narratives), and an ontology reconstruction task for objectively measuring event precision. We make the data, supplementary appendix, replication material, and visualizations of every historical episode available at a companion website www.crisisevents.org.

@@

If we could record every important interaction between countries in all of diplomacy, military conflict, and international political economy, how much unique information would this chronicle amount to, and how surprised would we be to see something new? In other words, what is the entropy of international relations? This record could in principle be unbounded, but the central conceit of social science is that there are structural regularities that limit what actors can do, their best options, and even which actors are likely to survive [@brecherInternationalStudiesTwentieth1999; @reiterShouldWeLeave2015]. If so, then these events can be systematically measured, and accordingly, massive effort is expended in social science attempting to record these regularities.^[See work on crises [@brecherCrisesWorldPolitics1982; @beardsleyInternationalCrisisBehavior2020], militarized disputes [@palmerMID5Dataset20112021; @giblerInternationalConflicts181620102018; @maozDyadicMilitarizedInterstate2019], wars [@sarkeesResortWar181620072010; @reiterRevisedLookInterstate2016], organized violence [@ralphsundbergUCDPGEDCodebook2016; @petterssonOrganizedViolence19892018], political violence [@raleighIntroducingACLEDArmed2010], sanctions [@felbermayrGlobalSanctionsData2020], trade [@barariDemocracyTradePolicy], and international agreements [@kinneDefenseCooperationAgreement2020; @owsiakInternationalBorderAgreements2018; @vabulasCooperationAutonomyBuilding2021], dispute resolution [@vabulasCooperationAutonomyBuilding2021; @frederickIssueCorrelatesWar2017], and diplomacy [@moyerWhatAreDrivers2020; @sechserMilitarizedCompellentThreats2011].] Thanks to improvements in natural language processing, more open-ended efforts have begun to capture entire unstructured streams of events from international news reports.^[See @liComprehensiveSurveySchemabased2021; @haltermanExtractingPoliticalEvents2020; @brandtPhoenixRealTimeEvent2018; @boscheeICEWSCodedEvent2015; @hegreIntroducingUCDPCandidate2020; @grantOUEventData2017. On event extraction from images and social media see @zhangCASMDeepLearningApproach2019 and @steinert-threlkeldFutureEventData2019.] How close these efforts are to accurately measuring all or even most of what is essential in international relations is an open empirical question, one for which we provide new evidence here.

Our contribution is a high coverage ontology and event dataset for key historical episodes in 20th and 21st-century international relations. We develop a large, flexible ontology of international events with the help of both human coders and natural language processing. We apply it sentence-by-sentence to an unusually high-quality corpus of historical narratives of international crises [@brecherInternationalStudiesTwentieth1999; @brecherCrisisEscalationWar2000; @wilkenfeldInterstateCrisesViolence2000; @jamesWhatWeKnow2019; @iakhnisCrisesWorldPolitics2019]. The result is a new lower bound estimate of how much actually happens between states during pivotal historical episodes. We then develop several methods for objectively gauging how well these event codings reconstruct the information contained in the original narrative. We conclude by benchmarking our event codings against several current state of the art event data collection efforts. We find that existing systems produce sequences of events that do not contain enough information to reconstruct the underlying historical episode. The underlying fine-grained variation in international affairs is unrecognizable through the lens of current quantification efforts.

This is a measurement paper that makes the following argument --- there is a real-world unobserved latent concept known as international relations, we propose a method for systematically measuring it, we successfully apply this method producing a new large-scale set of measurements, and those measurements exhibit several desirable kinds of internal and external validity, and those measurements out-perform other existing approaches. The article organizes that argument into eight sections: task definition; corpus; priors/existing state of the art; ICBe coding process; internal consistency; case study selection; recall; and precision. A final section concludes.

@@

We consider the measurement task of absstracting discrete events about a historical episode in international relations. The easiest way to convey the task is with an example. Figure 1 shows a narrative account of the Cuban Missile Crisis (1962) alongside a mapping from each natural language sentence to discrete machine-readable abstractive events. Formally, a historical episode, $H$, is demarcated by a period of time $[T_{start}, T_{end}] \in T$, a set of Players $p \in P$, and a set of behaviors they undertook during that time $b \in B$. International Relations, $IR$, is the system of regularities that govern the strategic interactions that world actors make during a historical episode, given their available options, preferences, beliefs, and expectations of choices made by others. We observe neither $H$ nor $IR$ directly. Rather the Historical Record, $HR$, produces documents $d \in D$ containing some relevant and true (as well as irrelevant and untrue) information about behaviors that were undertaken recorded in the form of unstructured natural language text. The task is to combine informative priors about $IR$ with an unstructured corpus $D$ to produce a series of structured discrete events, $e \in E$, that have high coverage, precision, and recall over what actually took place in history, $H$.

@@

temp <- ICBe_events_agreed_markdown %>%
        filter(crisno==196) %>%
        group_by(crisno, sentence_number_int_aligned) %>%
          filter(!duplicated(value_markdown %>% trimws() )) %>% #note there are dupe markdowns for some reason
        ungroup() %>%
        mutate(banding=as.numeric(as.factor(sentence_number_int_aligned)) %% 2)
temp$is_subsstring <- NA
for(i in 1:nrow(temp)){
  temp$is_subsstring[i] <-   stringr::str_detect(temp$value_markdown %>% tolower() %>% stringr::str_replace_all('[^a-z0-9]',"") ,
                                                                                pattern = stringr::fixed(temp$value_markdown[i] %>% tolower() %>% stringr::str_replace_all('[^a-z0-9]',"") )
                                                                              ) %>% replace_na(0) %>%  sum()
}
temp <-temp %>%
        mutate(value_markdown = ifelse(is_subsstring<=1,value_markdown,'')) %>%
        group_by(sentence_number_int_aligned) %>%
          filter(value_markdown!='' | length(value_markdown)==1) %>% # also added below to handle sentence 6
        ungroup() %>%
        mutate(crisno= ifelse(!is.na(lag(sentence_span_text)) & sentence_span_text==lag(sentence_span_text), "", crisno)) %>%
        mutate(sentence_number_int_aligned= ifelse(!is.na(lag(sentence_span_text)) &  sentence_span_text==lag(sentence_span_text), "", sentence_number_int_aligned)) %>%
        mutate(sentence_span_text= ifelse(!is.na(lag(sentence_span_text)) &  sentence_span_text==lag(sentence_span_text), "", sentence_span_text)) %>% 
            filter(value_markdown!='' | length(value_markdown)==1) 
      #C=crisno, 
      #head(100) %>%
ft_cuba_precision <- temp %>%
                     dplyr::select(S=sentence_number_int_aligned, ICB=sentence_span_text, ICBe=value_markdown)  %>%
                      flextable::as_flextable() %>%
        flextable::set_header_labels(values =
        list(
              S = "S",
              ICB = "Natural Language Sentences (ICB Corpus)",
              ICBe = "Machine-Readable Events (ICBe)"
             ) ) %>%

@@

Next, we draw informative prior beliefs about the underlying process of IR that we expect to govern behavior during historical episodes and their conversion to the historical record. We organize our prior beliefs along two overarching axes, summarized in detail by Table 1.

@@

We base our informed priors primarily on two sources of information. The first is the extensive existing measurement efforts of IR which we provide citations alongside each concept. Second, we performed preliminary natural language processing of the corpus and identified named entities and behaviors mentioned in the text. Verbs were matched to the most likely definition found in Wordnet [@millerWordNetLexicalDatabase1995], tallied, and then aggregated into a smaller number of hypernyms balancing conceptual detail and manageable sparsity for human coding (SI Appendix 1.2).

The second axis (columns) compares the very high ontological coverage of ICBe to existing state of the art systems in production and with global coverage. They begin with our contribution ICBe, alongside other event-level datasets including CAMEO dictionary lookup-based systems (Historical Phoenix [@althausClineCenterHistorical2019]; ICEWS [@boscheeICEWSCodedEvent2015; @hegreIntroducingUCDPCandidate2020]; Terrier [@grantOUEventData2017]), the Militarized Interstate Disputes Incidents dataset, and the UCDP-GED dataset [@ralphsundbergUCDPGEDCodebook2016; @petterssonOrganizedViolence19892018; @sundbergIntroducingUCDPGeoreferenced2013].^[Other datasets that insufficiently overlap ICBe's domain for comparison include BCOW [@lengMilitarizedInterstateCrises1988], WEIS [@mcclellandWorldEventInteraction1978], CREON [@hermannComparativeResearchEvents1984], CASCON [@bloomfieldCASCONIIIComputeraided1989], SHERFACS [@shermanSHERFACSCrossParadigmHierarchical2000], Real-Time Phoenix [@brandtPhoenixRealTimeEvent2018], and COfEE [@balaliCOfEEComprehensiveOntology2021] (see histories in @merrittMeasuringEventsInternational1994 and @schrodtTwentyYearsKansas2006).] The final set of columns compares episode-level datasets beginning with the original ICB project [@brecherInternationalCrisisBehavior; @brecherCrisesWorldPolitics1982; @beardsleyInternationalCrisisBehavior2020]; the Militarized Interstate Disputes dataset [@palmerMID5Dataset20112021; @giblerInternationalConflicts181620102018;@braithwaiteMIDLOCIntroducingMilitarized2010;@braithwaiteCodebookMilitarizedInterstate2009], and the Correlates of War [@sarkeesResortWar181620072010]. With the exception of large-scale CAMEO dictionary-based systems, the existing state of the art quantitative datasets ignore the vast majority of the information content found in international relations.^[See [@balaliCOfEEComprehensiveOntology2021] for a recent review of ontological depth and availability of Gold Standard example text.]

@@

The ICBe ontology follows a hierarchical design philosophy where a smaller number of significant decisions are made early on and then progressively refined into more specific details [@brustIntegratingDomainKnowledge2020].^[This process quickly focuses the coder on a smaller number of relevant options while also allowing them to apply multiple tags if the sentence explicitly includes more than one or there is insufficient evidence to choose only one tag. The guided coding process also allows for the possibility that earlier coarse decisions have less error than later fine-grained decisions.] Each coder was instructed to first thoroughly read the full crisis narrative and then presented with a custom graphical user interface (SI Appendix 2.1). Coders then proceeded sentence by sentence, choosing the number of events (0-3) that occurred, the highest behavior (thought, speech, or activity), a set of players ($P$), whether the means were primarily armed or unarmed, whether there was an increase or decrease in aggression (uncooperative/escalating or cooperative/de-escalating), and finally one or more non-mutually exclusive specific activities. Some additional details like location and timing information were always collected while other details were only collected if appropriate, e.g. force size, fatalities, domains, units, etc. A unique feature of the ontology is that thought, speech, and do behaviors can be nested into combinations, e.g. an offer for the U.S.S.R. to remove missiles from Cuba in exchange for the U.S. removing missiles from Turkey. Through compounding, the ontology can capture what players were said to have known, learned, or said about other specific fully described actions.

@@

We attribute the remaining disagreement primarily to three sources. First, we required coders to rate and justify their confidence in the coding. They reported low confidence for 20% of sentences; 45% of those were due to a mismatch between the ontology and the text ("survey doesn't fit event") and 46% were from a lack of information or confused writing in the source text (40% "more knowledge needed", 6% "confusing sentence"). Observed disagreement varied predictably with self-reported confidence (SI Appendix 2.4). Second, as intended, agreement is higher (75-80%) for questions with fewer options near the root of the ontology compared to agreement for questions near the leaves of the ontology (50%-60%). Third, individual coders exhibit nontrivial coding styles, e.g. some more expressive coders applied many tags per concept while others focused on only the single best match. We further observed unintended synonymity, e.g. the same information can be framed as either a threat to do something or a promise not to do something.

@@

The remaining two qualities we seek to measure are recall and precision of ICBe events in absolute terms and relative to other existing systems. We provide full ICB narratives, ICBe coding in an easy-to-read iconographic form, and a wide range of visualizations for every case on the companion website. In this paper, we focus on two deep case studies. The first is the Cuban Missile Crisis (Figure 1) which took place primarily in the second half of 1962, involved the United States, the Soviet Union, and Cuba, and is widely known for bringing the world to the brink of nuclear war (hereafter Cuban Missiles). The second is the Crimea-Donbas Crisis (SI Appendix 4.1) which took place primarily in 2014, involved Russia, Ukraine, and NATO, and within a decade spiraled into a full-scale invasion (hereafter Crimea-Donbas). Both cases involve a superpower in crisis with a neighbor, initiated by a change from a friendly to a hostile regime, with implications for economic and military security for the superpower, risked full-scale invasion, and eventually invited intervention by opposing superpowers. We choose these cases because they are substantively significant to 20th and 21st-century international relations, widely known across scientific disciplines and popular culture, and are sufficiently brief to evaluate in depth.

@@

Recall measures the share of desired information recovered by a sequence of coded events, $Pr(E|H)$, and is poorly defined for historical episodes. First, there is no genuine ground truth about what occurred, only surviving texts about it. Second, there is no \textit{a priori} guide to what information is necessary detail and what is ignorable trivia. History suffers from what is known as the Coastline Paradox [@mandelbrotFractalGeometryNature1983] --- it has a fractal dimension greater than one such that the more you zoom in, the more detail you will find about individual events and in between every two discrete events. The ICBe ontology is a proposal about what information is important, but we need an independent benchmark to evaluate whether that proposal is a good one and that allows for comparing proposals from event projects that had different goals. We need a yardstick for history.

Our strategy for dealing with both problems is a plausibly objective yardstick called a synthetic historical narrative. For both case studies, we collect a large diverse corpus of narratives spanning timelines, encyclopedia entries, journal articles, news reports, websites, and government documents. Using natural language processing (fully described in SI Appendix 3.1), we identify details that appear across multiple accounts. The more accounts that mention a detail, the more central it is to understanding the true historical episode. The theoretical motivation is that authors face word limits which force them to pick and choose which details to include, and they choose details that serve the specific context of the document they are producing. With a sufficiently large and diverse corpus of documents, we can vary the context while holding the overall episode constant and see which details tend to be invariant to context. Intuitively, a high-quality event dataset should have high recall for context invariant details both because of their broader relevance and also because they are easier to find in source material.

Synthetic historical narratives for Cuban Missiles (51 events drawn from 2020 documents) and Crimea-Donbas (30 events drawn from 971 documents) appear in Figure 2. Each row represents a detail that appeared in at least five documents along with an approximate start date, a handwritten summary, the number of documents it was mentioned in, and whether it could be identified in the text of our ICB corpus, in our ICBe events, or any of the competing systems.

From them, we draw several stylized facts. First, there is substantial variation in which details any one document will choose to include. Our ground truth ICB narratives contain 17/51 and 23/30 of the events from the synthetic narrative and include other details that are not in the synthetic narrative. Second, mentions of a detail across accounts are exponentially distributed with context-invariant details appearing dozens to hundreds of times more than context-dependent details. Third, crisis start and stop dates are arbitrary, and the historical record points to many precursor events as necessary detail for understanding later events, e.g. the U.S. was in a \textit{de facto} grey-scale war with Cuba before it invited Soviet military protection [@cormacGreyNewBlack2018] and Ukraine provided several security guarantees to Russia that were potentially undone, e.g. a long term lease on naval facilities in Crimea. Fourth, we find variation between the two cases. Cuban Missiles has a cleaner canonical end with the Soviets agreeing to withdraw missiles while Crimea-Donbas meekly ends with a second cease-fire agreement (Minsk II) but continued fighting. The canonical narrative of Cuban Missiles also includes high-level previously classified details, while the more recent Crimea-Donbas case reflects primarily public reporting.

We find substantive variation in recall across systems. Recall for each increases in the number of document mentions which is an important sign of validity for both them and our benchmark. The one outlier is Phoenix which is so noisy that it's flat to decreasing in mentions. The two episode-level datasets have very low coverage of contextual details. The two other dictionary systems ICEWs and Terrier have high coverage, with ICEWs outperforming Terrier. ICBe strictly dominates all of the systems but ICEWs in recall though we note that the small sample sizes mean these systems should be considered statistically indistinguishable. Importantly our corpus of ICB narratives has very high recall of frequently mentioned details giving us confidence in how those summaries were constructed, and ICBe lags only slightly behind showing that it left very little additional information on the table.

@@

      add_header_row(values = c("Cuban Missile Crisis (1962)", "", "Crimea-Donbas (2014)"), colwidths = c(kleft, 1,kright) ) %>%

@@

The other side of event measurement is precision, the degree to which a sequence of events correctly and usefully describes the information in history, $Pr(H|E)$. It does little good to recall a historical event but too vaguely (e.g. MIDs describes the Cuban Missile Crisis as a blockade, a show of force, and a stalemate) or with too much error (e.g. ICEWS records 263 "Detonate Nuclear Weapons" events between 1995-2019) to be useful for downstream applications. ICBe's ontology and coding system are designed to strike a balance so that the most important information is recovered accurately but also abstracted to a level that is still useful and interpretable. You should be able to lay out the events of a crisis on a timeline, as in Figure 3, and read off the macrostructure of an episode from each individual move. We call this visualization a crisis map, a directed graph intersected with a timeline, and provide crisis maps for every event dataset for each case study in SI Appendix 4.3 and 4.4 and for all crises on the companion website.

@@

We further want to verify individual event codings, which we can do in the case of ICBe because each event is mapped to a specific span of text. We develop the iconography system for presenting event codings as coherent statements that can be compared side by side to the original source narrative for Cuban Missiles (Figure 1), Crimea-Donbas (SI Appendix 4.1), and for every case on the companion website. We further provide a stratified sample of event codings alongside their source text (SI Appendix 4.2).
